Location and Scale-Invariant Power Transformations for Transforming Data to Normality

### 2025-04-12 manuscript_latex/manuscript.tex

\usepackage{graphicx}%
\usepackage{multirow}%
\usepackage{amsmath,amssymb,amsfonts}%
%% \usepackage{amsmath,amssymb,amsfonts}%
\usepackage{amsthm}%
\usepackage{mathrsfs}%
\usepackage[title]{appendix}%
\usepackage{xcolor}%
\usepackage{textcomp}%
\usepackage{manyfoot}%
\usepackage{booktabs}%
%% \usepackage{algorithm}%
%% \usepackage{algorithmicx}%
%% \usepackage{algpseudocode}%
\usepackage{listings}%
%%%%
\newtheorem*{definition}{Definition}
\DeclareMathOperator*{\argmax}{argmax}
\DeclareMathOperator*{\argmin}{argmin}
\DeclareMathOperator{\sgn}{sgn}

@@

For both invariant transformations, \(\lambda\), \(x_0\) and \(s\)
parameters can be obtained by maximising the log-likelihood function,
i.e.~using maximum likelihood estimation (MLE). A full derivation of the
log-likelihood function for both transformations is shown in Appendix A.
The location- and scale-invariant Box-Cox log-likelihood function is:

@@

\begin{table}
\begin{center}
\caption{Optimal weighting parameters and corresponding loss for location- and scale-invariant Box-Cox power transformations.
$p^{*}$ indicates use of the empirical distribution of feature values, $z$ the z-score of the transformed feature values,
and $r$ the residual error between the z-score of transformed feature values and the expected z-score according to the normal distribution.
The \textit{initial} column shows the starting parameter value for the optimisation process, with the corresponding boundary values in the \textit{limits} column. 
The \textit{optimal} column shows the optimal parameter values.
The \textit{loss} column shows the composite loss achieved by each method under optimised parameters.
This loss is based on residual errors of transformed features and deviations in $\lambda$ parameters
found compared to the $\lambda$ parameters found prior to inserting outliers.
The loss metric can be compared between different weighting methods.
}
\label{tab:optimal-weighting-parameters-box-cox}
\begin{tabular}{l r r r r r r r}

@@

\midrule
non-robust               & ---  & ---       & ---  & ---  & ---       & ---  & 49.6 \\
$p^{*}$ (step)           & 0.80 & $(0, 1]$  & 0.80 & ---  & ---       & ---  & 38.5 \\
$p^{*}$ (triangle)       & 0.80 & $(0, 1]$  & 0.01 & 1.00 & $(0, 1]$  & 0.86 & 43.7 \\
$p^{*}$ (tapered cosine) & 0.80 & $(0, 1]$  & 0.00 & 1.00 & $(0, 1]$  & 0.90 & 42.7 \\
$z$ (step)               & 1.28 & $(0, 10]$ & 2.39 & ---  & ---       & ---  & 52.2 \\
$z$ (triangle)           & 1.28 & $(0, 10]$ & 2.39 & 2.40 & $(0, 10]$ & 4.92 & 52.7 \\
$z$ (tapered cosine)     & 1.28 & $(0, 10]$ & 1.07 & 3.63 & $(0, 10]$ & 3.43 & 57.4 \\
$r$ (step)               & 0.50 & $(0, 10]$ & 0.83 & ---  & ---       & ---  & 53.1 \\
$r$ (triangle)           & 0.50 & $(0, 10]$ & 0.80 & 0.81 & $(0, 10]$ & 0.84 & 54.4 \\
$r$ (tapered cosine)     & 0.50 & $(0, 10]$ & 0.76 & 0.77 & $(0, 10]$ & 0.76 & 53.2 \\
\bottomrule
\end{tabular}
\end{center}
\end{table}

\begin{table}
\begin{center}
\caption{Optimal weighting parameters and corresponding loss for location- and scale-invariant Yeo-Johnson power transformations.
$p^{*}$ indicates use of the empirical distribution of feature values, $z$ the z-score of the transformed feature values,
and $r$ the residual error between the z-score of transformed feature values and the expected z-score according to the normal distribution. 
The \textit{initial} column shows the starting parameter value for the optimisation process, with the corresponding boundary values in the \textit{limits} column.
The \textit{loss} column shows the composite loss achieved by each method under optimised parameters.
This loss is based on residual errors of transformed features and deviations in $\lambda$ parameters
found compared to the $\lambda$ parameters found prior to inserting outliers.
The loss metric can be compared between different weighting methods.
}
\label{tab:optimal-weighting-parameters-yeo-johnson}
\begin{tabular}{l r r r r r r r}

@@

\midrule
non-robust               & ---  & ---       & ---  & ---  & ---       & ---  & 42.1 \\
$p^{*}$ (step)           & 0.80 & $(0, 1]$  & 0.78 & ---  & ---       & ---  & 35.1 \\
$p^{*}$ (triangle)       & 0.80 & $(0, 1]$  & 0.20 & 0.95 & $(0, 1]$  & 1.00 & 34.0 \\
$p^{*}$ (tapered cosine) & 0.80 & $(0, 1]$  & 0.54 & 0.95 & $(0, 1]$  & 1.00 & 32.6 \\
$z$ (step)               & 1.28 & $(0, 10]$ & 2.32 & ---  & ---       & ---  & 43.0 \\
$z$ (triangle)           & 1.28 & $(0, 10]$ & 1.28 & 1.96 & $(0, 10]$ & 3.43 & 49.7 \\
$z$ (tapered cosine)     & 1.28 & $(0, 10]$ & 0.41 & 1.96 & $(0, 10]$ & 3.75 & 51.3 \\
$r$ (step)               & 0.50 & $(0, 10]$ & 0.92 & ---  & ---       & ---  & 65.4 \\
$r$ (triangle)           & 0.50 & $(0, 10]$ & 0.87 & 1.00 & $(0, 10]$ & 0.87 & 66.0 \\
$r$ (tapered cosine)     & 0.50 & $(0, 10]$ & 1.06 & 1.00 & $(0, 10]$ & 1.07 & 67.1 \\
\bottomrule
\end{tabular}
\end{center}
\end{table}

@@

This results in nine types of power transformation. Transformation
parameters were optimised for each sequence. Subsequently the sum of
residual errors and sum of residual errors of the central portion
(\(\kappa = 0.80\)) were computed for each sequence after
transformation. Then, each method was ranked according to the sum of
residual errors (data without outliers) or the sum of residual errors of
the central portion (data with outliers). The average rank of each
transformation method was computed over all 10000 sequences in each
dataset. Average ranks for Yeo-Johnson transformations are shown in
Table \ref{tab:comparison_methods_simulation_yeo_johnson}. Location and
shift-invariant Yeo-Johnson transformation ranked best for all datasets
without outliers. The robust variant ranked best in dirty and shifted
datasets with outliers, but not in the clean dataset. Results for
Box-Cox transformations are shown in Table
\ref{tab:comparison_methods_simulations_box_cox}.

\begin{table}
\begin{center}
\caption{
Comparison of average rank between Yeo-Johnson transformation methods based on either residual error (without outliers) or residual error of the central portion
(with outliers; $\kappa = 0.80$) over 3 datasets with 10000 sequences each. The clean dataset consists of sequences derived through inverse Yeo-Johnson transformation
of data sampled from a standard normal distribution. The dirty dataset contains sequences sampled from asymmetric generalised normal distributions, centred at 0.
The shifted dataset also contains sequences sampled from asymmetric generalised normal distributions, but centred at 100, and scaled by 0.001.
Several transformation methods include normalisation before transformation, indicated by z-score normalisation (norm.) or robust scaling.
A rank of 1 is the best and a rank of 9 the worst. For each dataset, the best ranking transformation is marked in bold.
}
\label{tab:comparison_methods_simulations_yeo_johnson}
\begin{tabular}{l | l r r r r r r}

@@

\bottomrule
\end{tabular}
\end{center}
\end{table}

@@

Location- and scale-invariant power transformations are intended to
yield improved transformations to normality in the presence of large
shifts in location, distributions that due to location and scale are not
centered near zero, or both. Earlier, we assessed these transformations
using simulated data. In the following, they are evaluated using
examples from real datasets. We focus on the Yeo-Johnson transformation
because of its ability to handle features with negative values. Results
for Box-Cox transformations are shown in Appendix E.

@@

A common feature in health-related datasets is age. Here we use data on
228 patients with lung cancer that was collected and published by
Loprinzi et al. \citep{Loprinzi1994-cd}. The age in the cohort was
\(62.4 \pm 9.1\) (mean ± standard deviation) years. Applying
conventional and invariant Yeo-Johnson transformations to patient age
yielded the following results, see Figure
\ref{fig:experimental-results-invariance}: no transformation
(\(\sum r_i = 16.5\), \(p = 0.69\)); conventional transformation
(\(\lambda = 2.0\), \(\sum r_i = 11.5\), \(\mu_{YJ} = 1.8 \cdot 10^3\),
\(\sigma_{YJ} = 0.5 \cdot 10^3\), \(p = 0.96\)); Raymaekers and
Rousseeuw's robust adaptation (\(\lambda = 2.0\), \(\sum r_i = 11.5\),
\(\mu_{YJ} = 1.8 \cdot 10^3\), \(\sigma_{YJ} = 0.5 \cdot 10^3\),
\(p = 0.96\)); location- and scale-invariant transformation
(\(\lambda = 1.3\), \(\sum r_i = 8.8\), \(\mu_{YJ} = -1.2\),
\(\sigma_{YJ} = 1.1\), \(p = 0.98\)); and robust location- and
scale-invariant transformation (\(\lambda = 1.3\), \(\sum r_i = 9.3\),
\(\mu_{YJ} = -1.0\), \(\sigma_{YJ} = 1.1\), \(p = 0.93\)).

@@

Gorman, Williams and Fraser recorded body mass (in grams) of 342
penguins of three different species \citep{Gorman2014-eo}.
The body mass was \((4.2 \pm 0.8) \cdot 10^3\) (mean ± standard
deviation) grams, and not centrally normal (\(p < 0.001\)). Applying
conventional and invariant Yeo-Johnson transformations to body mass
yielded the following results, see Figure
\ref{fig:experimental-results-invariance}: no transformation (residual
sum \(\sum r_i = 48.0\), \(p < 0.001\)); conventional transformation
(\(\lambda = -0.5\), \(\sum r_i = 32.2\), \(\mu_{YJ} = 2.1\),
\(\sigma_{YJ} = 4 \cdot 10^{-3}\), \(p = 0.10\)); Raymaekers and
Rousseeuw's robust adaptation (\(\lambda = -0.5\), \(\sum r_i = 32.2\),
\(\mu_{YJ} = 2.1\), \(\sigma_{YJ} = 4 \cdot 10^{-3}\), \(p = 0.10\));
location- and scale-invariant transformation (\(\lambda = 0.5\),
\(\sum r_i = 26.8\), \(\mu_{YJ} = 0.9\), \(\sigma_{YJ} = 0.9\),
\(p = 0.28\)); and robust location- and scale-invariant transformation
(\(\lambda = 0.3\), \(\sum r_i = 22.0\), \(\mu_{YJ} = 0.7\),
\(\sigma_{YJ} = 0.9\), \(p = 0.69\)).

@@

Geospatial datasets usually contain coordinates. The Ames housing
dataset contains data on 2930 properties that were sold between 2006 and
2010 \citep{De-Cock2011-jf}, including their geospatial coordinates. The
latitude was \(42.03 \pm 0.02)\) (mean ± standard deviation). Applying
conventional and invariant Yeo-Johnson transformations to latitude
yielded the following results, see Figure
\ref{fig:experimental-results-invariance}: no transformation (residual
sum \(\sum r_i = 328\), \(p < 0.001\)); conventional transformation
(\(\lambda = 62.1\), \(\sum r_i = 319\),
\(\mu_{YJ} = 4.8 \cdot 10^{99}\), \(\sigma_{YJ} = 0.1 \cdot 10^{99}\),
\(p < 0.001\)); Raymaekers and Rousseeuw's robust adaptation
(\(\lambda = 95.4\), \(\sum r_i = 315\),
\(\mu_{YJ} = 6.4 \cdot 10^{153}\), \(\sigma_{YJ} = 0.3 \cdot 10^{153}\),
\(p < 0.001\)); location- and scale-invariant transformation
(\(\lambda = 1.5\), \(\sum r_i = 326\), \(\mu_{YJ} = -1.2\),
\(\sigma_{YJ} = 0.8\), \(p < 0.001\)); and robust location- and
scale-invariant transformation (\(\lambda = 1.1\), \(\sum r_i = 308\),
\(\mu_{YJ} = -1.3\), \(\sigma_{YJ} = 1.2\), \(p < 0.001\)).

@@

\begin{itemize}
\tightlist
\item
  very easy: \(\text{AUC} \geq 0.90\) or \(\text{RRSE} \leq 0.10\) (57
  datasets)
\item
  easy: \(0.90 > \text{AUC} \geq 0.80\) or
  \(0.30 \geq \text{RRSE} > 0.10\) (40 datasets)
\item
  intermediate: \(0.80 > \text{AUC} \geq 0.70\) or
  \(0.60 \geq \text{RRSE} > 0.30\) (52 datasets)
\item
  difficult: \(0.70 > \text{AUC} \geq 0.60\) or
  \(0.80 \geq \text{RRSE} > 0.60\) (33 datasets)
\item
  very difficult: \(0.60 > \text{AUC} \geq 0.50\) or
  \(1.00 \geq \text{RRSE} > 0.80\) (48 datasets)
\item
  unsolvable: \(\text{AUC} < 0.50\) or \(\text{RRSE} > 1.00\) (1
  dataset)
\end{itemize}

@@

This work has the following limitation: We observed several numerical
stability issues for optimisation criteria other than MLE (Appendix B).
These appear in regions where transformation parameters would lead to
very large or small numbers when using conventional power
transformations. For MLE stability issues were not observed.

@@

\section{Appendix A: Log-likelihood functions for location- and
scale-invariant power
transformation}\label{appendix-a}

@@

\section{Appendix B: Optimisation of transformation
parameters}\label{appendix-b}

@@

\subsubsection{Cramér-von Mises
criterion}\label{cramer-von-mises-criterion}

The Cramér-von Mises criterion is also based on the empirical
distribution function of \(\mathbf{X}\). We define the Cramér-von Mises
criterion as follows:

@@

Here \(w_i\) are weights, and \(\mu\) and \(\sigma\) are location and
scale parameters. For non-robust power transformations, all \(w_i = 1\).
The criterion is similar to the Cramér-von Mises test statistic 
\citep{Cramer1928-rc, Von_Mises1928-ef}, aside from a additive scalar value and the introduction of
weights. This criterion, like the Anderson-Darling criterion, seeks to
minimise the squared Euclidean distance between the EDF and the normal
CDF. Unlike the Anderson-Darling criterion, this criterion weights all
instances equally.

For conventional power transformations with a fixed shift parameter, the
transformation \(\phi^{\lambda, x_0, s} (\mathbf{X})\) may be
substituted by \(\phi^{\lambda} (\mathbf{X})\) in the definition of the
Cramér-von Mises criterion.

@@

Invariance of location- and scale-invariant power transformations was
assessed using the optimisation criteria in Appendix B. 
This follows the simulation in the main manuscript, where MLE was
used for optimization. In short, we first randomly drew \(10000\) values
from a normal distribution:
\(\mathbf{X}_{\text{normal}} = \left\{x_1, x_2, \ldots, x_{10000} \right\} \sim \mathcal{N}\left(0, 1\right)\),
or equivalently
\(\mathbf{X}_{\text{normal}} = \left\{x_1, x_2, \ldots, x_{10000} \right\} \sim \mathcal{AGN}\left(0, 1/\sqrt{2}, 0.5, 2\right)\).
The second distribution was a right-skewed normal distribution
\(\mathbf{X}_{\text{right}} = \left\{x_1, x_2, \ldots, x_{10000} \right\} \sim \mathcal{AGN}\left(0, 1/\sqrt{2}, 0.2, 2\right)\).
The third distribution was a left-skewed normal distribution
\(\mathbf{X}_{\text{left}} = \left\{x_1, x_2, \ldots, x_{10000} \right\} \sim \mathcal{AGN}\left(0, 1/\sqrt{2}, 0.8, 2\right)\).

@@

\section{Appendix C: Central normality test and empirical central
normality test critical
values}\label{appendix-c}

@@

\begin{table}
\begin{center}
\caption{Critical values of test statistic $\tau_{\alpha, n, \kappa = 0.80}$ for
\textbf{central normality} at $\kappa = 0.80$, as a function of significance 
level $\alpha$ and number of instances $n$. The values shown must be divided by $100$.}
\label{tab:central-normality-critical-statistic}
\begin{tabular}{l | r r r r r r r r r}

@@

\midrule
    5 & 210.79 & 95.75 & 66.27 & 49.69 & 35.93 & 27.60 & 19.29 & 13.41 & 11.00 \\
   10 &  48.04 & 31.44 & 26.96 & 23.91 & 21.11 & 18.16 & 13.66 & 10.14 &  8.67 \\
   20 &  29.13 & 22.13 & 19.75 & 17.94 & 16.00 & 13.81 & 10.54 &  8.18 &  7.18 \\
   50 &  17.31 & 14.01 & 12.63 & 11.48 & 10.28 &  9.00 &  7.00 &  5.54 &  4.94 \\
  100 &  12.19 & 10.04 &  9.02 &  8.21 &  7.38 &  6.47 &  5.04 &  3.99 &  3.56 \\
  200 &   8.68 &  7.11 &  6.40 &  5.83 &  5.22 &  4.57 &  3.59 &  2.86 &  2.56 \\ 
  500 &   5.61 &  4.54 &  4.08 &  3.70 &  3.33 &  2.91 &  2.28 &  1.82 &  1.63 \\
 1000 &   3.93 &  3.19 &  2.87 &  2.60 &  2.33 &  2.05 &  1.61 &  1.29 &  1.15 \\
 2000 &   2.71 &  2.24 &  2.02 &  1.84 &  1.66 &  1.46 &  1.14 &  0.91 &  0.82 \\
 5000 &   1.73 &  1.42 &  1.28 &  1.17 &  1.05 &  0.92 &  0.72 &  0.58 &  0.52 \\
10000 &   1.22 &  1.00 &  0.90 &  0.82 &  0.74 &  0.65 &  0.51 &  0.41 &  0.37 \\
\bottomrule
\end{tabular}
\end{center}
\end{table}

>>>>

\usepackage{graphicx}%
\usepackage{multirow}%
\usepackage{amsmath,amssymb,amsfonts}%
%% \usepackage{amsmath,amssymb,amsfonts}%
\usepackage{amsthm}%
\usepackage{mathrsfs}%
\usepackage[title]{appendix}%
\usepackage{xcolor}%
\usepackage{textcomp}%
\usepackage{manyfoot}%
\usepackage{booktabs}%
%% \usepackage{algorithm}%
%% \usepackage{algorithmicx}%
%% \usepackage{algpseudocode}%
\usepackage{listings}%
\usepackage{placeins}
%%%%
\newtheorem*{definition}{Definition}
\DeclareMathOperator*{\argmax}{argmax}
\DeclareMathOperator*{\argmin}{argmin}
\DeclareMathOperator{\sgn}{sgn}

@@

For both invariant transformations, \(\lambda\), \(x_0\) and \(s\)
parameters can be obtained by maximising the log-likelihood function,
i.e.~using maximum likelihood estimation (MLE). A full derivation of the
log-likelihood function for both transformations is shown in \hyperref[appendix-a]{Appendix A}.
The location- and scale-invariant Box-Cox log-likelihood function is:

@@

\begin{table}
\caption{Optimal weighting parameters and corresponding loss for location- and scale-invariant Box-Cox power transformations.
$p^{*}$ indicates use of the empirical distribution of feature values, $z$ the z-score of the transformed feature values,
and $r$ the residual error between the z-score of transformed feature values and the expected z-score according to the normal distribution.
The \textit{initial} column shows the starting parameter value for the optimisation process, with the corresponding boundary values in the \textit{limits} column. 
The \textit{optimal} column shows the optimal parameter values.
The \textit{loss} column shows the composite loss achieved by each method under optimised parameters.
This loss is based on residual errors of transformed features and deviations in $\lambda$ parameters
found compared to the $\lambda$ parameters found prior to inserting outliers.
The loss metric can be compared between different weighting methods.
}
\label{tab:optimal-weighting-parameters-box-cox}
\begin{tabular}{l r r r r r r r}

@@

\midrule
non-robust               & ---  & ---       & ---  & ---  & ---       & ---  & 49.6 \\
$p^{*}$ (step)           & 0.80 & $(0, 1]$  & 0.80 & ---  & ---       & ---  & 38.5 \\
$p^{*}$ (triangle)       & 0.80 & $(0, 1]$  & 0.01 & 1.00 & $(0, 1]$  & 0.86 & 43.7 \\
$p^{*}$ (tapered cosine) & 0.80 & $(0, 1]$  & 0.00 & 1.00 & $(0, 1]$  & 0.90 & 42.7 \\
$z$ (step)               & 1.28 & $(0, 10]$ & 2.39 & ---  & ---       & ---  & 52.2 \\
$z$ (triangle)           & 1.28 & $(0, 10]$ & 2.39 & 2.40 & $(0, 10]$ & 4.92 & 52.7 \\
$z$ (tapered cosine)     & 1.28 & $(0, 10]$ & 1.07 & 3.63 & $(0, 10]$ & 3.43 & 57.4 \\
$r$ (step)               & 0.50 & $(0, 10]$ & 0.83 & ---  & ---       & ---  & 53.1 \\
$r$ (triangle)           & 0.50 & $(0, 10]$ & 0.80 & 0.81 & $(0, 10]$ & 0.84 & 54.4 \\
$r$ (tapered cosine)     & 0.50 & $(0, 10]$ & 0.76 & 0.77 & $(0, 10]$ & 0.76 & 53.2 \\
\bottomrule
\end{tabular}
\end{table}

\begin{table}
\caption{Optimal weighting parameters and corresponding loss for location- and scale-invariant Yeo-Johnson power transformations.
$p^{*}$ indicates use of the empirical distribution of feature values, $z$ the z-score of the transformed feature values,
and $r$ the residual error between the z-score of transformed feature values and the expected z-score according to the normal distribution. 
The \textit{initial} column shows the starting parameter value for the optimisation process, with the corresponding boundary values in the \textit{limits} column.
The \textit{loss} column shows the composite loss achieved by each method under optimised parameters.
This loss is based on residual errors of transformed features and deviations in $\lambda$ parameters
found compared to the $\lambda$ parameters found prior to inserting outliers.
The loss metric can be compared between different weighting methods.
}
\label{tab:optimal-weighting-parameters-yeo-johnson}
\begin{tabular}{l r r r r r r r}

@@

\midrule
non-robust               & ---  & ---       & ---  & ---  & ---       & ---  & 42.1 \\
$p^{*}$ (step)           & 0.80 & $(0, 1]$  & 0.78 & ---  & ---       & ---  & 35.1 \\
$p^{*}$ (triangle)       & 0.80 & $(0, 1]$  & 0.20 & 0.95 & $(0, 1]$  & 1.00 & 34.0 \\
$p^{*}$ (tapered cosine) & 0.80 & $(0, 1]$  & 0.54 & 0.95 & $(0, 1]$  & 1.00 & 32.6 \\
$z$ (step)               & 1.28 & $(0, 10]$ & 2.32 & ---  & ---       & ---  & 43.0 \\
$z$ (triangle)           & 1.28 & $(0, 10]$ & 1.28 & 1.96 & $(0, 10]$ & 3.43 & 49.7 \\
$z$ (tapered cosine)     & 1.28 & $(0, 10]$ & 0.41 & 1.96 & $(0, 10]$ & 3.75 & 51.3 \\
$r$ (step)               & 0.50 & $(0, 10]$ & 0.92 & ---  & ---       & ---  & 65.4 \\
$r$ (triangle)           & 0.50 & $(0, 10]$ & 0.87 & 1.00 & $(0, 10]$ & 0.87 & 66.0 \\
$r$ (tapered cosine)     & 0.50 & $(0, 10]$ & 1.06 & 1.00 & $(0, 10]$ & 1.07 & 67.1 \\
\bottomrule
\end{tabular}
\end{table}

@@

This results in nine types of power transformation. Transformation
parameters were optimised for each sequence. Subsequently the sum of
residual errors and sum of residual errors of the central portion
(\(\kappa = 0.80\)) were computed for each sequence after
transformation. Then, each method was ranked according to the sum of
residual errors (data without outliers) or the sum of residual errors of
the central portion (data with outliers). The average rank of each
transformation method was computed over all 10000 sequences in each
dataset. Average ranks for Yeo-Johnson transformations are shown in
Table \ref{tab:comparison_methods_simulations_yeo_johnson}. Location and
shift-invariant Yeo-Johnson transformation ranked best for all datasets
without outliers. The robust variant ranked best in dirty and shifted
datasets with outliers, but not in the clean dataset. Results for
Box-Cox transformations are shown in Table
\ref{tab:comparison_methods_simulations_box_cox}.

\begin{table}
\caption{
Comparison of average rank between Yeo-Johnson transformation methods based on either residual error (without outliers) or residual error of the central portion
(with outliers; $\kappa = 0.80$) over 3 datasets with 10000 sequences each. The clean dataset consists of sequences derived through inverse Yeo-Johnson transformation
of data sampled from a standard normal distribution. The dirty dataset contains sequences sampled from asymmetric generalised normal distributions, centred at 0.
The shifted dataset also contains sequences sampled from asymmetric generalised normal distributions, but centred at 100, and scaled by 0.001.
Several transformation methods include normalisation before transformation, indicated by z-score normalisation (norm.) or robust scaling.
A rank of 1 is the best and a rank of 9 the worst. For each dataset, the best ranking transformation is marked in bold.
}
\label{tab:comparison_methods_simulations_yeo_johnson}
\begin{tabular}{l | l r r r r r r}

@@

\bottomrule
\end{tabular}
\end{table}

@@

Location- and scale-invariant power transformations are intended to
yield improved transformations to normality in the presence of large
shifts in location, distributions that due to location and scale are not
centered near zero, or both. Earlier, we assessed these transformations
using simulated data. In the following, they are evaluated using
examples from real datasets. We focus on the Yeo-Johnson transformation
because of its ability to handle features with negative values. Results
for Box-Cox transformations are shown in \hyperref[appendix-e]{Appendix E}.

@@

A common feature in health-related datasets is age. Here we use data on
228 patients with lung cancer that was collected and published by
Loprinzi et al. \citep{Loprinzi1994-cd}. The age in the cohort was
\(62.4 \pm 9.1\) (mean $\pm$ standard deviation) years. Applying
conventional and invariant Yeo-Johnson transformations to patient age
yielded the following results, see Figure
\ref{fig:experimental-results-invariance}: no transformation
(\(\sum r_i = 16.5\), \(p = 0.69\)); conventional transformation
(\(\lambda = 2.0\), \(\sum r_i = 11.5\), \(\mu_{YJ} = 1.8 \cdot 10^3\),
\(\sigma_{YJ} = 0.5 \cdot 10^3\), \(p = 0.96\)); Raymaekers and
Rousseeuw's robust adaptation (\(\lambda = 2.0\), \(\sum r_i = 11.5\),
\(\mu_{YJ} = 1.8 \cdot 10^3\), \(\sigma_{YJ} = 0.5 \cdot 10^3\),
\(p = 0.96\)); location- and scale-invariant transformation
(\(\lambda = 1.3\), \(\sum r_i = 8.8\), \(\mu_{YJ} = -1.2\),
\(\sigma_{YJ} = 1.1\), \(p = 0.98\)); and robust location- and
scale-invariant transformation (\(\lambda = 1.3\), \(\sum r_i = 9.3\),
\(\mu_{YJ} = -1.0\), \(\sigma_{YJ} = 1.1\), \(p = 0.93\)).

@@

Gorman, Williams and Fraser recorded body mass (in grams) of 342
penguins of three different species \citep{Gorman2014-eo}.
The body mass was \((4.2 \pm 0.8) \cdot 10^3\) (mean $\pm$ standard
deviation) grams, and not centrally normal (\(p < 0.001\)). Applying
conventional and invariant Yeo-Johnson transformations to body mass
yielded the following results, see Figure
\ref{fig:experimental-results-invariance}: no transformation (residual
sum \(\sum r_i = 48.0\), \(p < 0.001\)); conventional transformation
(\(\lambda = -0.5\), \(\sum r_i = 32.2\), \(\mu_{YJ} = 2.1\),
\(\sigma_{YJ} = 4 \cdot 10^{-3}\), \(p = 0.10\)); Raymaekers and
Rousseeuw's robust adaptation (\(\lambda = -0.5\), \(\sum r_i = 32.2\),
\(\mu_{YJ} = 2.1\), \(\sigma_{YJ} = 4 \cdot 10^{-3}\), \(p = 0.10\));
location- and scale-invariant transformation (\(\lambda = 0.5\),
\(\sum r_i = 26.8\), \(\mu_{YJ} = 0.9\), \(\sigma_{YJ} = 0.9\),
\(p = 0.28\)); and robust location- and scale-invariant transformation
(\(\lambda = 0.3\), \(\sum r_i = 22.0\), \(\mu_{YJ} = 0.7\),
\(\sigma_{YJ} = 0.9\), \(p = 0.69\)).

@@

Geospatial datasets usually contain coordinates. The Ames housing
dataset contains data on 2930 properties that were sold between 2006 and
2010 \citep{De-Cock2011-jf}, including their geospatial coordinates. The
latitude was \(42.03 \pm 0.02)\) (mean $\pm$ standard deviation). Applying
conventional and invariant Yeo-Johnson transformations to latitude
yielded the following results, see Figure
\ref{fig:experimental-results-invariance}: no transformation (residual
sum \(\sum r_i = 328\), \(p < 0.001\)); conventional transformation
(\(\lambda = 62.1\), \(\sum r_i = 319\),
\(\mu_{YJ} = 4.8 \cdot 10^{99}\), \(\sigma_{YJ} = 0.1 \cdot 10^{99}\),
\(p < 0.001\)); Raymaekers and Rousseeuw's robust adaptation
(\(\lambda = 95.4\), \(\sum r_i = 315\),
\(\mu_{YJ} = 6.4 \cdot 10^{153}\), \(\sigma_{YJ} = 0.3 \cdot 10^{153}\),
\(p < 0.001\)); location- and scale-invariant transformation
(\(\lambda = 1.5\), \(\sum r_i = 326\), \(\mu_{YJ} = -1.2\),
\(\sigma_{YJ} = 0.8\), \(p < 0.001\)); and robust location- and
scale-invariant transformation (\(\lambda = 1.1\), \(\sum r_i = 308\),
\(\mu_{YJ} = -1.3\), \(\sigma_{YJ} = 1.2\), \(p < 0.001\)).

@@

\begin{itemize}
\item
  very easy: \(\text{AUC} \geq 0.90\) or \(\text{RRSE} \leq 0.10\) (57
  datasets)
\item
  easy: \(0.90 > \text{AUC} \geq 0.80\) or
  \(0.30 \geq \text{RRSE} > 0.10\) (40 datasets)
\item
  intermediate: \(0.80 > \text{AUC} \geq 0.70\) or
  \(0.60 \geq \text{RRSE} > 0.30\) (52 datasets)
\item
  difficult: \(0.70 > \text{AUC} \geq 0.60\) or
  \(0.80 \geq \text{RRSE} > 0.60\) (33 datasets)
\item
  very difficult: \(0.60 > \text{AUC} \geq 0.50\) or
  \(1.00 \geq \text{RRSE} > 0.80\) (48 datasets)
\item
  unsolvable: \(\text{AUC} < 0.50\) or \(\text{RRSE} > 1.00\) (1
  dataset)
\end{itemize}

@@

This work has the following limitation: We observed several numerical
stability issues for optimisation criteria other than MLE (\hyperref[appendix-b]{Appendix B}).
These appear in regions where transformation parameters would lead to
very large or small numbers when using conventional power
transformations. For MLE stability issues were not observed.

@@

\section{Log-likelihood functions for location- and
scale-invariant power
transformation}\label{appendix-a}

@@

\section{Optimisation of transformation
parameters}\label{appendix-b}

@@

\subsubsection{Cram\'{e}r-von Mises
criterion}\label{cramer-von-mises-criterion}

The Cram\'{e}r-von Mises criterion is also based on the empirical
distribution function of \(\mathbf{X}\). We define the Cram\'{e}r-von Mises
criterion as follows:

@@

Here \(w_i\) are weights, and \(\mu\) and \(\sigma\) are location and
scale parameters. For non-robust power transformations, all \(w_i = 1\).
The criterion is similar to the Cram\'{e}r-von Mises test statistic 
\citep{Cramer1928-rc, Von_Mises1928-ef}, aside from a additive scalar value and the introduction of
weights. This criterion, like the Anderson-Darling criterion, seeks to
minimise the squared Euclidean distance between the EDF and the normal
CDF. Unlike the Anderson-Darling criterion, this criterion weights all
instances equally.

For conventional power transformations with a fixed shift parameter, the
transformation \(\phi^{\lambda, x_0, s} (\mathbf{X})\) may be
substituted by \(\phi^{\lambda} (\mathbf{X})\) in the definition of the
Cram\'{e}r-von Mises criterion.

@@

Invariance of location- and scale-invariant power transformations was
assessed using the optimisation criteria in \hyperref[appendix-b]{Appendix B}. 
This follows the simulation in the main manuscript, where MLE was
used for optimization. In short, we first randomly drew \(10000\) values
from a normal distribution:
\(\mathbf{X}_{\text{normal}} = \left\{x_1, x_2, \ldots, x_{10000} \right\} \sim \mathcal{N}\left(0, 1\right)\),
or equivalently
\(\mathbf{X}_{\text{normal}} = \left\{x_1, x_2, \ldots, x_{10000} \right\} \sim \mathcal{AGN}\left(0, 1/\sqrt{2}, 0.5, 2\right)\).
The second distribution was a right-skewed normal distribution
\(\mathbf{X}_{\text{right}} = \left\{x_1, x_2, \ldots, x_{10000} \right\} \sim \mathcal{AGN}\left(0, 1/\sqrt{2}, 0.2, 2\right)\).
The third distribution was a left-skewed normal distribution
\(\mathbf{X}_{\text{left}} = \left\{x_1, x_2, \ldots, x_{10000} \right\} \sim \mathcal{AGN}\left(0, 1/\sqrt{2}, 0.8, 2\right)\).

@@

\FloatBarrier

\section{Central normality test and empirical central
normality test critical
values}\label{appendix-c}

@@

\begin{table}
\caption{Critical values of test statistic $\tau_{\alpha, n, \kappa = 0.80}$ for
\textbf{central normality} at $\kappa = 0.80$, as a function of significance 
level $\alpha$ and number of instances $n$. The values shown must be divided by $100$.}
\label{tab:central-normality-critical-statistic}
\begin{tabular}{l | r r r r r r r r r}

@@

\midrule
    5 & 210.79 & 95.75 & 66.27 & 49.69 & 35.93 & 27.60 & 19.29 & 13.41 & 11.00 \\
   10 &  48.04 & 31.44 & 26.96 & 23.91 & 21.11 & 18.16 & 13.66 & 10.14 &  8.67 \\
   20 &  29.13 & 22.13 & 19.75 & 17.94 & 16.00 & 13.81 & 10.54 &  8.18 &  7.18 \\
   50 &  17.31 & 14.01 & 12.63 & 11.48 & 10.28 &  9.00 &  7.00 &  5.54 &  4.94 \\
  100 &  12.19 & 10.04 &  9.02 &  8.21 &  7.38 &  6.47 &  5.04 &  3.99 &  3.56 \\
  200 &   8.68 &  7.11 &  6.40 &  5.83 &  5.22 &  4.57 &  3.59 &  2.86 &  2.56 \\ 
  500 &   5.61 &  4.54 &  4.08 &  3.70 &  3.33 &  2.91 &  2.28 &  1.82 &  1.63 \\
 1000 &   3.93 &  3.19 &  2.87 &  2.60 &  2.33 &  2.05 &  1.61 &  1.29 &  1.15 \\
 2000 &   2.71 &  2.24 &  2.02 &  1.84 &  1.66 &  1.46 &  1.14 &  0.91 &  0.82 \\
 5000 &   1.73 &  1.42 &  1.28 &  1.17 &  1.05 &  0.92 &  0.72 &  0.58 &  0.52 \\
10000 &   1.22 &  1.00 &  0.90 &  0.82 &  0.74 &  0.65 &  0.51 &  0.41 &  0.37 \\
\bottomrule
\end{tabular}
\end{table}
